---
title: "Linear regression: Sonadora temperature and elevation example"
author: "Mila Matich"
date: "2026-05-15"
format:
  pdf:
    toc: true # use this to display a table of contents
execute:
  message: false # use this to make sure messages don't show up
  warning: false # use this to make sure warnings don't show up
---

# Set up

```{r}
#| label: libraries-and-data

library(tidyverse) # general use
library(janitor) # cleaning data frames
library(here) # file/folder organization
library(ggeffects) # generating model predictions
library(flextable) # creating tables

# temperature data from Ramirez 2024
sonadora <- read_csv(here("data", "Temp_SonadoraGradient_Daily.csv"))
```

# Sonadora temperature example

Data from Ramirez, A. 2024. Sonadora elevational plots: long-term monitoring of air temperature ver 877108. Environmental Data Initiative. https://doi.org/10.6073/pasta/6b66eecae3092d8f2340b5132dec38ab (Accessed 2025-05-14).

## a. Questions and hypotheses

Question: Does elevation (in meters) predict temperature (in °C)?

H~0~: Elevation (m) does not predict temperature (°C).

H~A~: Elevation (m) predicts temperature (°C).

## b. Cleaning and summarizing

```{r}
#| label: temperature-cleaning

# creating new clean data frame
sonadora_clean <- sonadora |> 
  # clean column names
  clean_names() |> 
  # make the data frame longer
  pivot_longer(cols = plot_250:plot_1000,
               names_to = "plot_name",
               values_to = "temp_c") |> 
  # separate plot name from elevation
  separate_wider_delim(cols = plot_name,
                       delim = "_",
                       names = c("plot", "elevation_m"),
                       cols_remove = FALSE) |> 
  # remove plot column
  select(-plot) |> 
  # make sure elevation is read as a number
  mutate(elevation_m = as.numeric(elevation_m))

# summarizing
sonadora_sum <- sonadora_clean |> 
  # group by plot and elevation
  group_by(plot_name, elevation_m) |> 
  # calculate mean temperature at each elevation
  summarize(mean_temp_c = mean(temp_c, na.rm = TRUE)) |> 
  # undo the group_by function
  ungroup() |> 
  # arrange in order of elevation
  arrange(elevation_m)
```

## c. Exploratory data visualization

```{r}
#| label: temperature-visualization

# base layer: ggplot
ggplot(data = sonadora_sum, 
       # labeling x and y-axis
       aes(x = elevation_m,
           y = mean_temp_c)) +
  # first layer: points representing temperature at each elevation
  geom_point(color = "dodgerblue3") # adding costum color

```

## d. Temperature model

```{r}
#| label: temperature-model

# model
temperature_model <- lm(
  mean_temp_c ~ elevation_m, # formula: response ~ predictor
  data = sonadora_sum # data frame
  )

```

### Diagnostics

```{r}
#| label: temperature-diagnostics

# diagnostics
par(mfrow = c(2, 2)) # displays plot in 2x2 grid !
plot(temperature_model) # display diagnostic plot 

```

### Model summary

```{r}
#| label: temperature-model-summary

# displaying model summary 
summary(temperature_model)

```

**For each meter of elevation gain, you would expect a decrease in temperature equivalent to 0.01 ± 0.0003 (SE) °C.**

### Generating predictions

```{r}
#| label: temperature-model-predictions

# model predictions
temperature_preds <- ggpredict(
  temperature_model,      # model object
  terms = "elevation_m"   # predictor (in quotation marks)
)

# calculate the temperature prediction at elevation = 900
ggpredict(
  temperature_model,           # model object
  terms = "elevation_m[900]"   # predictor (in quotation marks) and predictor value in brackets
)
```

**At 900 m, the predicted temperature is 19.8 °C (95% CI: [19.6, 20.0]).**

### Visualizing model predictions

```{r}
#| label: temperature-mode-prediction-visualization

# base layer
ggplot(data = sonadora_sum, 
       # labeling x and y axes 
       aes(x = elevation_m, # x = elevation 
           y = mean_temp_c)) + # y = mean temp 
  # first layer: temperature at each elevation
  geom_point(color = "pink1") + # 
  # 95% CI ribbon
  # uses model prediction data frame
  geom_ribbon(data = temperature_preds,
              aes(x = x,
                  y = predicted,
                  ymin = conf.low,
                  ymax = conf.high),
              color = "pink",
              fill = "lightyellow",
              alpha = 0.2) +
  # model prediction line
  # uses model prediction data frame
  geom_line(data = temperature_preds,
              aes(x = x,
                  y = predicted)) +
  # axis labels
  labs(x = "Elevation (m)",
       y = "Mean temperature (\U00B0 C)") +
  theme_bw()
```

### Creating a table with model coefficients, 95% confidence intervals, and more

```{r}
#| label: temperature-model-table

as_flextable(temperature_model)
```

**END OF SONADORA EXAMPLE**
